---
title: "OPIOIDUSE"
output: html_document
---
# Project name : studying opioid test therapy requirements data 

# Project purpose : Cleaning and creating a table with transgender data 

# Code author name : Mahankali Sravani

# Date last edited : 3/6/2022

# Location of data used : St.Louis, MO

####################################################################################################

## reading the opioid usage file 

```{r}
library(readr)
opioid.use <- read.csv("pdmp_2017_kff_ch4.csv")
```

## loading the tidyverse module 

```{r}
library(package = "tidyverse")
```

## studying the summary of the data

```{r}
summary(object = opioid.use)
```
Observation: All the variables are characters with length 51

## selecting the variable OPioid.Step.Therapy.Requirements, converting it to numeric, and recoding the    variable

```{r}
set.seed(seed = 30)


opioid.use.cleaned <- opioid.use%>%select(Opioid.Step.Therapy.Requirements) %>%
  mutate(Opioid.Step.Therapy.Requirements = as.numeric(recode(.x =Opioid.Step.Therapy.Requirements,
                                  `No` = 0L,
                                  `Yes` = 1L)))

```

## studying the summary of the new created variable opioid.use.cleaned 

```{r}
summary(object = opioid.use.cleaned)
```
Observation: the data is converted into numeric data

## examining the oioid.use.cleaned

```{r}
view(opioid.use.cleaned)
```

Observation: the data is converted into 1s and 0s, 0 for No and 1 for Yes.

1. FIRST QUESTION

## Conerting the data intlo the data frame 

```{r}
Df2 <- data.frame(
  response=sample(opioid.use.cleaned,51,replace=T),
  stringsAsFactors=F)
```

## Converting the data frame into a table of Opioid.Step.Therapy.Requrements

```{r}
opioid.policy.kff  <- table(Df2$response.Opioid.Step.Therapy.Requirements)

opioid.policy.kff 
```
Observation: 12 states didn't have Opioid.Step.Therapy.Requirements,and 39 states opted for them.

## calculating the proportion of the opioid.policy.kff using prop.table() function

```{r}
prop.table(opioid.policy.kff)
```
Observation: 23.5% of states didn't opt for Opioid,Step.Therapy.Requirements and 76.47% of states opted for it.

2. SECOND QUESTION

## using dbinorm to find the probability

```{r}
dbinom(x = 5, size = 10, prob = .765)
```
Observation: If 10 states were chosen at random, there is a 0.047 probability that exactly 5 of them 
             would have adopted the step therapy policy.

3. THIRD QUESTION

## using dbinorm to find the probability

```{r}
dbinom(x = 5, size = 15, prob = .765)
```
Observation: If 15 states were chosen at random, there is a 0.000404 probability that exactly 5 of                 them would have adopted the step therapy policy.

4. FOURTH QUESTION

## selecting sample of 30 states with seed value and recoding it. 

```{r}
set.seed(seed = 50)
# sample 30 states at random
opioid.use.cleaned1 <- opioid.use%>%select(Opioid.Step.Therapy.Requirements) %>%
  mutate(Opioid.Step.Therapy.Requirements = as.numeric(recode(.x =Opioid.Step.Therapy.Requirements,
                                  `No` = 0,
                                  `Yes` = 1)))%>%

sample_n(size = 30)
```

## coverting into data frame

```{r}
Df4 <- data.frame(
  response=sample(opioid.use.cleaned1,51,replace=T),
  stringsAsFactors=F)
opioid.policy.kff1  <- table(Df4$response.Opioid.Step.Therapy.Requirements)
opioid.policy.kff1 
```
Observation: Out of 30 sample states, 7 of them didn't opt for the step therapy and 23 of them opted               for it

## studying the proportions using prop.table() function.

```{r}
prop.table(opioid.policy.kff1)
```
Observation: Observation: 23.33% of states didn't opt for Opioid,Step.Therapy.Requirements and 76.66%              of states opted for it

## selecting the reponses and assigning it to the variable requirement

```{r}
requirement <- Df4$response.Opioid.Step.Therapy.Requirements
```

## calculating 95%  confidence  interval  for  the proportion. 

```{r}
opioid.policy.kff.Df4 <- Df4 %>%
summarize(mean.s1 = mean(x = requirement),
     se.s1 = sd(x = requirement)/sqrt(x = length(x = requirement)),
     lower.ci.s1 = mean.s1 - 1.96 * se.s1,
     upper.ci.s1 = mean.s1 + 1.96 * se.s1)

opioid.policy.kff.Df4
```
Observation: The mean of the sample is 0.766, standard error is 0.0785, lower class interval is 0.612
             and the upper class interval is 0.92060
-------------------------------------------------------------------------------------------------
5. FIFTH QUESTION

## reading the opioid dist to needle exchange data and assigning it to opioid.dist.each

```{r}
opioid.dist.exch <- read.csv("opioid_dist_to_needle_exchange_2018.csv")
```

## renaming the variable VALUE in the data as distance

```{r}
dist.exch.cleaned <- opioid.dist.exch %>%
rename('distance' = VALUE)
```

## calculating mean and standard deviation of the data

```{r}
dist.exch.cleaned %>%
summarize(mean.dist = mean(x = distance),
sd.dist = sd(x = distance),
n = n())
```
Observation: mean of the data is 105.48, and standard deviation is 87.35

6. SIXTH QUESTION

## viewing the data dist.exch.cleaned

```{r}
view(dist.exch.cleaned)
```

## visualizing the data 

```{r}
dist.exch.cleaned %>%
ggplot(aes(x = distance)) +
geom_histogram(fill = "#D776AF", color = "white", bins = 30) +
theme_minimal() +
labs(x = "Miles to needle exchange",
y = "Number of counties")
```
Observation: the data is left skewed, and we need to transform the data.

## the data is transformed into cuberoot, log, inverse, and square root function

```{r}
dist.exch.cleaned <- dist.exch.cleaned%>%
mutate(miles.cube.root = distance^(1/3)) %>%
mutate(miles.log = log(x = distance)) %>%
mutate(miles.inverse = 1/distance) %>%
mutate(miles.sqrt = sqrt(x = distance))

```

## visualizing the log transformation of the data.

```{r}
logged <- dist.exch.cleaned %>%
ggplot(aes(x = miles.log)) +
geom_histogram(fill = "#D776AF", color = "white", bins = 30) +
theme_minimal() +
labs(x = "Log of miles to needle exchange", y = "")
```

## visualizing the inverse transformation of the data

```{r}
inversed <- dist.exch.cleaned %>%
ggplot(aes(x = miles.inverse)) +
geom_histogram(fill = "#D776AF", color = "white", bins = 30) +
theme_minimal() + xlim(0, 1) +
labs(x = "Inverse of miles to needle exchange", y = "Number of counties")
```

## visualizing the square root transformation of the data

```{r}
squareroot <- dist.exch.cleaned %>%
ggplot(aes(x = miles.sqrt)) +
geom_histogram(fill = "#D776AF", color = "white", bins = 30) +
theme_minimal() +
labs(x = "square root of miles to needle exchange", y = "")
```

## visualzing the cube root transformation of the data

```{r}
cuberoot <- dist.exch.cleaned %>%
ggplot(aes(x = miles.cube.root)) +
geom_histogram(fill = "#D776AF", color = "white", bins = 30) +
theme_minimal() +
labs(x = "Cube root of miles to needle exchange", y = "Number of counties")
```

## printing the graphs in a grid

```{r}
gridExtra::grid.arrange(cuberoot, logged, inversed, squareroot)
```
Observation: the cuberoot transformation is the perfect transformation to get normalized curve

## calculating the mean and standard deviation of the transformed data

```{r}
dist.exch.cleaned %>%
summarize(mean.tran.dist = mean(x = miles.cube.root),
       sd.tran.dist = sd(x = miles.cube.root))
```
Observation: the mean is 4.349 and the standard deviation is 1.327

## calculating St. Louis County distance  to needle exchange

```{r}
st.louis.distance = 12.27^(1/3)
st.louis.distance
```
Observation: the needle exchange is 2.306 miles from St.Louis County

## calculating St. Louis County  z-score  for  distance  to needle exchange

```{r}
z.score <- (st.louis.distance - 4.349566)/1.327338	
z.score
```
Observation: negative value for a county implies distance was lower than the mean. And in this case                z-score is -1.539 

7. SEVENTH QUESTION

## Selecting 500 countries from sample at random

```{r}
set.seed(seed = 18)
# sample 500 counties at random
counties.500 <- dist.exch.cleaned %>%
sample_n(size = 500, replace = TRUE)
```

## calculating the mean, standard deviation and standard error for the sample

```{r}
counties.500 %>%
summarize(mean.dist = mean(x = distance),
    sd.dist = sd(x = distance),
 se.dist = sd(x = distance)/sqrt(x = length(x = distance)))
```
Observation: the mean of the sample is 106.6, standard deviation is 84.89, and standard error is 3.79

8.EIGTH QUESTION

## Calculating the mean of population data

```{r}
dist.exch.cleaned %>%
summarize(mean.dist = mean(x = distance))

```
## calculating mean of sample 

```{r}
counties.500 %>%
summarize(mean.dist = mean(x = distance))
```
Observation: The mean of the population and the sample are almost similar most of the sample means are              within one or two standard deviations of the population mean.

9. NINTH QUESTION

## calculating 95% confidence intervals of the sample

```{r}
counties.500 <- counties.500 %>%
  summarize(mean.s1 = mean(x = distance),
     se.s1 = sd(x = distance)/sqrt(x = length(x = distance)),
     lower.ci.s1 = mean.s1 - 1.96 * se.s1,
     upper.ci.s1 = mean.s1 + 1.96 * se.s1)

counties.500
```
Observation: The mean, standard error, lower class interval and upper class interval are 106.6, 3.79, 99.16, and 114.04 respectively.

####################################################################################################
